# Song Length Word Count Since 1964
```{r setopts, echo=FALSE, warning=FALSE, message=FALSE}
library("knitr")
library("XML")
library("ggplot2")
library("stringr")
library("gridExtra")
opts_chunk$set(echo=FALSE, warning=FALSE, fig.height=4)
```

The songs used for analysis were scraped from Wikipedia’s entry for each year. The lyrics were scraped from [songlyrics.com](http://www.songlyrics.com/) (~98%) or manually entered (~2%) from another source, most frequently [metrolyrics.com](http://www.metrolyrics.com/), if they were not available on [songlyrics.com](http://www.songlyrics.com/). See this post with the R code here.

```{r library_and_data}
library("ggplot2")

allsongs <- read.csv("final_all_songs.csv", stringsAsFactors=FALSE) 
```
```{r wordcount}
for ( i in 1:length(allsongs$Lyrics) ) { 
     allsongs$wordcount[i] <- length( strsplit(allsongs$Lyrics[i], " ")[[1]] )
}

countbyyear <- aggregate(wordcount ~ Year, data = allsongs, FUN = mean) 
```

```{r wordcount_plot}
ggplot(countbyyear, aes(Year, wordcount)) + geom_line(aes(color="persong"),size=1.25) + labs(title="Average Number of Words Per Song\n in Billboard Year-End Hot 100 Songs\n 1964-2014", x="",y="") + coord_cartesian(ylim=c(0,500)) + theme(legend.position="none", plot.title = element_text(size=14, face="bold", vjust=2), axis.ticks.y = element_blank(), panel.background = element_rect(fill = 'white'))
```

Potential complication - "repeat chorus 3x" or [chorus] in lyrics instead of actually writing it out

The doubling is due to longer songs - from 2.5 to 4 minutes since the 1960s [[1]](http://www.wired.com/2014/07/why-are-songs-on-the-radio-about-the-same-length/).  

And an increased proportion of Year-End Hot 100 songs featuring more than one artist. 

```{r feat_artists}
feat <- allsongs[grep("[Ff]eaturing|[Ff]eat|[Dd]uet|[Ww]ith", allsongs$Artist), ]

featbyyear <- aggregate(Artist ~ Year, data = feat, FUN = length)
```

```{r feat_artists_plot}
ggplot(featbyyear, aes(as.numeric(Year), Artist)) + geom_line(aes(color="artists"),size=1.25) + labs(title="Songs Featuring 2+ Artists in Billboard Year-End Hot 100 Songs\n 1964-2014", x="",y="") + coord_cartesian(ylim=c(0,100)) + theme(legend.position="none", plot.title = element_text(size=14, face="bold", vjust=2), axis.ticks.y = element_blank(), panel.background = element_rect(fill = 'white'))
```
